Reproducibility of COVID-19 pre-prints

### 2021-12-08 outputs/paper/paper.Rmd
Update references

---
title: "Reproducibility of COVID-19 pre-prints"
author: 
  - Annie Collins^[University of Toronto.]
  - Rohan Alexander^[University of Toronto, rohan.alexander@utoronto.ca.]
thanks: "**Acknowledgements:** We thank Amy Farrow, Jessica Gronsbell, Monica Alexander, and Thomas William Rosenthal for helpful comments. **Funding statement:** Collins thanks CANSSI Ontario and Monica Alexander for financial support. **Data accessibility statement:** Code and data are available at: https://github.com/anniecollins/reproducibility_markers_in_covid19_preprints. **Author contributions:** Collins had the original idea and wrote the first draft of the paper. Collins and Alexander obtained and created the datasets. Collins and Alexander analysed the data. All authors interpreted the data and contributed writing the paper and approved the final version."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "To examine the reproducibility of COVID-19 research, we create a dataset of pre-prints posted to arXiv, bioRxiv, medRxiv, and SocArXiv between 28 January 2020 and 30 June 2021 that are related to COVID-19. We extract the text from these pre-prints and parse them looking for keyword markers signalling the availability of the data and code underpinning the pre-print. For the pre-prints that are in our sample, we are unable to find markers of either open data or open code for 75 per cent of those on arXiv, 67 per cent of those on bioRxiv, 79 per cent of those on medRxiv, and 85 per cent of those on SocArXiv. We conclude that there may be value in having authors categorize the degree of openness of their pre-print as part of the pre-print submissions process, and more broadly, there is a need to better integrate open science training into a wide range of fields."
output:
  bookdown::pdf_document2:
    keep_tex:  true
toc: FALSE
bibliography: references.bib
---

@@

Open data is generally accepted to be beneficial to the scientific process and to a paper's reproducibility potential, hence it is concerning that around 75 per cent of pre-prints in our sample contained no open data markers. This concern is slightly mitigated by recognition of challenges in working with biomedical data compared with data in other fields, notably privacy and ethics concerns when working with personal data [@Floca2014]. The COVID-19 pandemic has seen open science initiatives, as evidenced by the creation of open data repositories such as the dashboard maintained by the Center for Systems Science and Engineering at Johns Hopkins University [@citeCSSE] or the large number of publishers who have removed paywalls from published COVID-19 research [@gill2020]. While the intention at the start of the pandemic was that there would be 'clear statements regarding the availability of underlying data' [@citeWellcome] some retractions of work have been based on 'unreliable or nonexistent data' [@da2021optimizing]. SocArXiv asks submitters to make explicit the degree of openness in the data, and this is something that could be easily adopted by other pre-print repositories.

@@

In the context of the above factors, it was encouraging to find in our analysis that the proportion of pre-prints with open data or code posted to arXiv increased from 7 per cent pre-pandemic to 25 per cent for COVID-19-related pre-prints. This pattern, however, was not observed among the analyzed bioRxiv and medRxiv pre-prints, and may just reflect the nature of COVID-19 pre-prints. With many pre-prints from these repositories still pertaining to epidemiological modelling, one might hope that they should universally be subject to the same analysis as conducted by @jin2020 as for the examples above, which is made possible by the availability of relevant code and data. Our analysis suggests a need for future investigation and potential overall improvement in open science standards for these types of pre-prints (subject to the data and code considerations already discussed). This need is again emphasized by the new-found speed at which pre-prints may gain public, media, and political attention in the context of the pandemic, particularly those from medRxiv and bioRxiv.

@@

Beyond pre-prints, COVID-19 has influenced publication and peer review processes, with timelines for COVID-19 papers being expedited at the expense of longer waits for other scientific research [@else2020]. It is important that open data and code standards be maintained in published work as well. In our sample, published pre-prints contain open data or code markers in similar proportions to their unpublished counterparts, a pattern that was present for pre-prints related to COVID-19 and those posted in 2019. This appears initially to alleviate some concerns over the relationship between open data and publication bias, that is, the potential that journals have favoured novel yet less transparent or reproducible papers over those with null results but a high standard of open science practices. However, publication bias is complex and this result should be approached with caution. Concerns have already been raised through systemic reviews of COVID-19 publications [@raynaud2021], and oversights in data accessibility have led to high profile retractions of publications in the past; for example, papers from *The Lancet* and the *New England Journal of Medicine* which were withdrawn due to concerns over the private nature of their underlying dataset [@ledford2020].

>>>>

---
title: "Reproducibility of COVID-19 pre-prints"
author: 
  - Annie Collins^[University of Toronto.]
  - Rohan Alexander^[University of Toronto, rohan.alexander@utoronto.ca.]
thanks: "**Preprint:** This present manuscript is a version of a preprint by the authors [@collins2021reproducibility]. The text, tables and figures of this manuscript overlap with those of the preprint version by the authors. **Acknowledgements:** We thank Amy Farrow, Jessica Gronsbell, Monica Alexander, and Thomas William Rosenthal for helpful comments. **Funding statement:** Collins thanks CANSSI Ontario and Monica Alexander for financial support. **Data accessibility statement:** Code and data are available at: https://github.com/anniecollins/reproducibility_markers_in_covid19_preprints. **Author contributions:** Collins had the original idea and wrote the first draft of the paper. Collins and Alexander obtained and created the datasets. Collins and Alexander analysed the data. All authors interpreted the data and contributed writing the paper and approved the final version."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "To examine the reproducibility of COVID-19 research, we create a dataset of pre-prints posted to arXiv, bioRxiv, medRxiv, and SocArXiv between 28 January 2020 and 30 June 2021 that are related to COVID-19. We extract the text from these pre-prints and parse them looking for keyword markers signalling the availability of the data and code underpinning the pre-print. For the pre-prints that are in our sample, we are unable to find markers of either open data or open code for 75 per cent of those on arXiv, 67 per cent of those on bioRxiv, 79 per cent of those on medRxiv, and 85 per cent of those on SocArXiv. We conclude that there may be value in having authors categorize the degree of openness of their pre-print as part of the pre-print submissions process, and more broadly, there is a need to better integrate open science training into a wide range of fields."
output:
  bookdown::pdf_document2:
    keep_tex:  true
toc: FALSE
bibliography: references.bib
---

@@

Open data is generally accepted to be beneficial to the scientific process and to a paper's reproducibility potential, hence it is concerning that around 75 per cent of pre-prints in our sample contained no open data markers. This concern is slightly mitigated by recognition of challenges in working with biomedical data compared with data in other fields, notably privacy and ethics concerns when working with personal data [@Floca2014]. The COVID-19 pandemic has seen open science initiatives, as evidenced by the creation of open data repositories such as the dashboard maintained by the Center for Systems Science and Engineering at Johns Hopkins University [@citeCSSE] or the large number of publishers who have removed paywalls from published COVID-19 research [@gill2020]. While the intention at the start of the pandemic was that there would be 'clear statements regarding the availability of underlying data' [@citeWellcome] some retractions of work have been based on 'unreliable or nonexistent data' [@da2021optimizing]. SocArXiv asks submitters to make explicit the degree of openness in the data, and this is something that could be adopted by other pre-print repositories.

@@

In the context of the above factors, it was encouraging to find in our analysis that the proportion of pre-prints with open data or code posted to arXiv increased from 7 per cent pre-pandemic to 25 per cent for COVID-19-related pre-prints. This pattern, however, was not observed among the analyzed bioRxiv and medRxiv pre-prints, and may just reflect the nature of COVID-19 pre-prints. With many pre-prints from these repositories still pertaining to epidemiological modelling, one might hope that they should universally be subject to the same analysis as conducted by @jin2020 as for the examples above, which is made possible by the availability of relevant code and data. Our analysis suggests a need for future investigation and potential overall improvement in open science standards for these types of pre-prints (subject to the data and code considerations already discussed). This need is again emphasized by the new-found speed at which pre-prints may gain public, media, and political attention in the context of the pandemic, particularly those from medRxiv and bioRxiv. One further concern is raised by @da2020silently, who show that there are pre-prints on those two pre-print servers---medRxiv and bioRxiv---that were withdrawn or retracted with relatively little information about the underlying reason, after gaining substantial media attention.

@@

Beyond pre-prints, COVID-19 has influenced publication and peer review processes, with timelines for COVID-19 papers being expedited at the expense of longer waits for other scientific research [@else2020]. It is important that open data and code standards be maintained in published work as well. In our sample, published pre-prints contain open data or code markers in similar proportions to their unpublished counterparts, a pattern that was present for pre-prints related to COVID-19 and those posted in 2019. This appears initially to alleviate some concerns over the relationship between open data and publication bias, that is, the potential that journals have favoured novel yet less transparent or reproducible papers over those with null results but a high standard of open science practices. However, publication bias is complex and this result should be approached with caution. Concerns have already been raised through systemic reviews of COVID-19 publications [@raynaud2021], and oversights in data accessibility have led to high profile retractions of publications in the past; for example, papers from *The Lancet* and the *New England Journal of Medicine* which were withdrawn due to concerns over the private nature of their underlying dataset [@ledford2020]. Further, @cabanac2021day show that not all pre-prints are linked to their subsequent peer-reviewed publication, which may further bias our results. And @oikonomidi2020changes and @Beroe051821 show that differences between updated versions of the same pre-print can be substantial; again this is something that we do not account for and could bias our results.

### 2021-12-08 outputs/paper/paper.tex
Update references

\title{Reproducibility of COVID-19 pre-prints\thanks{\textbf{Acknowledgements:} We thank Amy Farrow, Jessica Gronsbell, Monica Alexander, and Thomas William Rosenthal for helpful comments. \textbf{Funding statement:} Collins thanks CANSSI Ontario and Monica Alexander for financial support. \textbf{Data accessibility statement:} Code and data are available at: \url{https://github.com/anniecollins/reproducibility_markers_in_covid19_preprints}. \textbf{Author contributions:} Collins had the original idea and wrote the first draft of the paper. Collins and Alexander obtained and created the datasets. Collins and Alexander analysed the data. All authors interpreted the data and contributed writing the paper and approved the final version.}}
\author{Annie Collins\footnote{University of Toronto.} \and Rohan Alexander\footnote{University of Toronto, \href{mailto:rohan.alexander@utoronto.ca}{\nolinkurl{rohan.alexander@utoronto.ca}}.}}
\date{01 October 2021}

@@

Open data is generally accepted to be beneficial to the scientific process and to a paper's reproducibility potential, hence it is concerning that around 75 per cent of pre-prints in our sample contained no open data markers. This concern is slightly mitigated by recognition of challenges in working with biomedical data compared with data in other fields, notably privacy and ethics concerns when working with personal data (Floca 2014). The COVID-19 pandemic has seen open science initiatives, as evidenced by the creation of open data repositories such as the dashboard maintained by the Center for Systems Science and Engineering at Johns Hopkins University (Dong, Du, and Gardner 2020) or the large number of publishers who have removed paywalls from published COVID-19 research (Gill 2020). While the intention at the start of the pandemic was that there would be `clear statements regarding the availability of underlying data' (Wellcome Trust 2020) some retractions of work have been based on `unreliable or nonexistent data' (Teixeira da Silva, Bornemann-Cimenti, and Tsigaris 2021). SocArXiv asks submitters to make explicit the degree of openness in the data, and this is something that could be easily adopted by other pre-print repositories.

@@

In the context of the above factors, it was encouraging to find in our analysis that the proportion of pre-prints with open data or code posted to arXiv increased from 7 per cent pre-pandemic to 25 per cent for COVID-19-related pre-prints. This pattern, however, was not observed among the analyzed bioRxiv and medRxiv pre-prints, and may just reflect the nature of COVID-19 pre-prints. With many pre-prints from these repositories still pertaining to epidemiological modelling, one might hope that they should universally be subject to the same analysis as conducted by Jin et al. (2020) as for the examples above, which is made possible by the availability of relevant code and data. Our analysis suggests a need for future investigation and potential overall improvement in open science standards for these types of pre-prints (subject to the data and code considerations already discussed). This need is again emphasized by the new-found speed at which pre-prints may gain public, media, and political attention in the context of the pandemic, particularly those from medRxiv and bioRxiv.

@@

Beyond pre-prints, COVID-19 has influenced publication and peer review processes, with timelines for COVID-19 papers being expedited at the expense of longer waits for other scientific research (Else 2020). It is important that open data and code standards be maintained in published work as well. In our sample, published pre-prints contain open data or code markers in similar proportions to their unpublished counterparts, a pattern that was present for pre-prints related to COVID-19 and those posted in 2019. This appears initially to alleviate some concerns over the relationship between open data and publication bias, that is, the potential that journals have favoured novel yet less transparent or reproducible papers over those with null results but a high standard of open science practices. However, publication bias is complex and this result should be approached with caution. Concerns have already been raised through systemic reviews of COVID-19 publications (Raynaud et al. 2020), and oversights in data accessibility have led to high profile retractions of publications in the past; for example, papers from \emph{The Lancet} and the \emph{New England Journal of Medicine} which were withdrawn due to concerns over the private nature of their underlying dataset (Ledford and Noorden 2020).

@@

@@

@@

>>>>

\title{Reproducibility of COVID-19 pre-prints\thanks{\textbf{Preprint:} This present manuscript is a version of a preprint by the authors (Collins and Alexander 2021). The text, tables and figures of this manuscript overlap with those of the preprint version by the authors. \textbf{Acknowledgements:} We thank Amy Farrow, Jessica Gronsbell, Monica Alexander, and Thomas William Rosenthal for helpful comments. \textbf{Funding statement:} Collins thanks CANSSI Ontario and Monica Alexander for financial support. \textbf{Data accessibility statement:} Code and data are available at: \url{https://github.com/anniecollins/reproducibility_markers_in_covid19_preprints}. \textbf{Author contributions:} Collins had the original idea and wrote the first draft of the paper. Collins and Alexander obtained and created the datasets. Collins and Alexander analysed the data. All authors interpreted the data and contributed writing the paper and approved the final version.}}
\author{Annie Collins\footnote{University of Toronto.} \and Rohan Alexander\footnote{University of Toronto, \href{mailto:rohan.alexander@utoronto.ca}{\nolinkurl{rohan.alexander@utoronto.ca}}.}}
\date{08 December 2021}

@@

Open data is generally accepted to be beneficial to the scientific process and to a paper's reproducibility potential, hence it is concerning that around 75 per cent of pre-prints in our sample contained no open data markers. This concern is slightly mitigated by recognition of challenges in working with biomedical data compared with data in other fields, notably privacy and ethics concerns when working with personal data (Floca 2014). The COVID-19 pandemic has seen open science initiatives, as evidenced by the creation of open data repositories such as the dashboard maintained by the Center for Systems Science and Engineering at Johns Hopkins University (Dong, Du, and Gardner 2020) or the large number of publishers who have removed paywalls from published COVID-19 research (Gill 2020). While the intention at the start of the pandemic was that there would be `clear statements regarding the availability of underlying data' (Wellcome Trust 2020) some retractions of work have been based on `unreliable or nonexistent data' (Teixeira da Silva, Bornemann-Cimenti, and Tsigaris 2021). SocArXiv asks submitters to make explicit the degree of openness in the data, and this is something that could be adopted by other pre-print repositories.

@@

In the context of the above factors, it was encouraging to find in our analysis that the proportion of pre-prints with open data or code posted to arXiv increased from 7 per cent pre-pandemic to 25 per cent for COVID-19-related pre-prints. This pattern, however, was not observed among the analyzed bioRxiv and medRxiv pre-prints, and may just reflect the nature of COVID-19 pre-prints. With many pre-prints from these repositories still pertaining to epidemiological modelling, one might hope that they should universally be subject to the same analysis as conducted by Jin et al. (2020) as for the examples above, which is made possible by the availability of relevant code and data. Our analysis suggests a need for future investigation and potential overall improvement in open science standards for these types of pre-prints (subject to the data and code considerations already discussed). This need is again emphasized by the new-found speed at which pre-prints may gain public, media, and political attention in the context of the pandemic, particularly those from medRxiv and bioRxiv. One further concern is raised by Teixeira da Silva (2020), who show that there are pre-prints on those two pre-print servers---medRxiv and bioRxiv---that were withdrawn or retracted with relatively little information about the underlying reason, after gaining substantial media attention.

@@

Beyond pre-prints, COVID-19 has influenced publication and peer review processes, with timelines for COVID-19 papers being expedited at the expense of longer waits for other scientific research (Else 2020). It is important that open data and code standards be maintained in published work as well. In our sample, published pre-prints contain open data or code markers in similar proportions to their unpublished counterparts, a pattern that was present for pre-prints related to COVID-19 and those posted in 2019. This appears initially to alleviate some concerns over the relationship between open data and publication bias, that is, the potential that journals have favoured novel yet less transparent or reproducible papers over those with null results but a high standard of open science practices. However, publication bias is complex and this result should be approached with caution. Concerns have already been raised through systemic reviews of COVID-19 publications (Raynaud et al. 2020), and oversights in data accessibility have led to high profile retractions of publications in the past; for example, papers from \emph{The Lancet} and the \emph{New England Journal of Medicine} which were withdrawn due to concerns over the private nature of their underlying dataset (Ledford and Noorden 2020). Further, Cabanac, Oikonomidi, and Boutron (2021) show that not all pre-prints are linked to their subsequent peer-reviewed publication, which may further bias our results. And Oikonomidi et al. (2020) and Bero et al. (2021) show that differences between updated versions of the same pre-print can be substantial; again this is something that we do not account for and could bias our results.

@@

\leavevmode\hypertarget{ref-Beroe051821}{}%
Bero, Lisa, Rosa Lawrence, Louis Leslie, Kellia Chiu, Sally McDonald, Matthew J Page, Quinn Grundy, et al. 2021. {``Cross-Sectional Study of Preprints and Final Journal Publications from COVID-19 Studies: Discrepancies in Results Reporting and Spin in Interpretation.''} \emph{BMJ Open} 11 (7). \url{https://doi.org/10.1136/bmjopen-2021-051821}.

\leavevmode\hypertarget{ref-cabanac2021day}{}%
Cabanac, Guillaume, Theodora Oikonomidi, and Isabelle Boutron. 2021. {``Day-to-Day Discovery of Preprint--Publication Links.''} \emph{Scientometrics} 126 (6): 5285--5304.

\leavevmode\hypertarget{ref-collins2021reproducibility}{}%
Collins, Annie, and Rohan Alexander. 2021. {``Reproducibility of COVID-19 Pre-Prints.''} \url{http://arxiv.org/abs/2107.10724}.

@@

\leavevmode\hypertarget{ref-oikonomidi2020changes}{}%
Oikonomidi, Theodora, Isabelle Boutron, Olivier Pierre, Guillaume Cabanac, and Philippe Ravaud. 2020. {``Changes in Evidence for Studies Assessing Interventions for COVID-19 Reported in Preprints: Meta-Research Study.''} \emph{BMC Medicine} 18 (1): 1--10.

@@

\leavevmode\hypertarget{ref-da2020silently}{}%
Teixeira da Silva, Jaime A. 2020. {``Silently Withdrawn or Retracted Preprints Related to Covid-19 Are a Scholarly Threat and a Potential Public Health Risk: Theoretical Arguments and Suggested Recommendations.''} \emph{Online Information Review}.
